---
title: "Image Compression Project"
output: html_document
date: "2026-09-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(readr)
img1 <- read_csv("image1.csv")
img1 <- data.matrix(img1)
rank_k_pca_approx <- function(img, k) {
  
  n <- nrow(img)
  p <- ncol(img)
  k <- min(k, n, p) 
  
  pca <- prcomp(img, center = TRUE, scale. = FALSE)
  
  scores_k   <- pca$x[, 1:k, drop = FALSE]        
  loadings_k <- pca$rotation[, 1:k, drop = FALSE]   
  
  approx <- scores_k %*% t(loadings_k)            
  approx <- sweep(approx, 2, pca$center, "+")        
  
  error <- norm(img - approx, type = "F")
  
  list(approximation = approx, error = error)
}

result <- rank_k_pca_approx(img1, k = 10)
dim(result$approximation) 
image(result$approximation, col = gray.colors(256))
plot(result$approximation)

```

## IMAGE 1 

**Here is the function that will plot the original image and its rank approximations for the first image. This function also plots the rank k approximation value against the error for that particular approximation.**

```{r, echo=FALSE}
img1 <- read_csv("image1.csv")

img_iterations <- function(ex_image){
  ex_image <- data.matrix(ex_image)
  results_vector <-c()
  for (i in 1:10){
    result <- rank_k_pca_approx(ex_image, k = i)
    plot(result$approximation, xlab = paste("X for k =", i), ylab = "Y")

    results_vector<-append(results_vector, result$error)
  }
  k_values <- c(1,2,3,4,5,6,7,8,9,10)
  plot(k_values, results_vector, xlab = "K", ylab = "Error term")
    
}
img_iterations(img1)



  
```
**For the first image, I think that the closes approximation is probably from the k = 7 iteration, as it is the lowest k values to begin to most resemble the original image.**

**Below is the plot of the first eigenvector, followed by a plot of the first PC score **
```{r, echo = FALSE}
eigen_plot <- function(ex_image){
pca <- prcomp(data.matrix(ex_image), center = TRUE, scale. = FALSE)
ev1 <- pca$rotation[, 1]   

plot(ev1, type = "b", xlab = "Index", ylab = "Loadings",
     main = "First Eigenvector (PC1 Loadings)")
scores <- pca$x
plot(scores[,1], type = "b", xlab = "Index", ylab = "PC1 score",
     main = "PC1 scores plotted")
}

eigen_plot(img1)
```
##Interpretation of PC1 scores and Eigenvector plots

**For the PC1 score plot, each number represents how far a feature projects in the new direction. In terms of the Eigenvectors plot, each index is how much a particular variable contributes to the principal component itself.**

##Image 2

**Here are the rank approximations for the second image, as well as the plot of error terms and k values.**
```{r, echo = FALSE}
img2 <- read_csv("image2.csv")

img_iterations(img2)
```
**For this image, all the approximations seem to be close. We could even choose something as small as k = 3, which does not look different at all from the original image plot.**

**Below is the plot of the first eigenvector, followed by a plot of the first PC score.**
```{r, echo=FALSE}
eigen_plot(img2)

```

**Similar to the first image, the PC scores furthest away from zero are the features that are most compatible with the new "pattern" the PCA algorithm determined. For eigenvectors, the scores determine how much a feature contributes to the principal component itself.**

##Image 3

**Here are the rank approximations for the third image, as well as the plot of error terms and k values.**
```{r, echo=FALSE}
img3 <- read_csv("image3.csv")
img_iterations(img3)
```

**For this "best approximation" of the original image, I will choose k to equal 8. It resembles the shape and data clustering the closest.**

**Below are the plots of the PC1 scores and the eigenvectors.**

```{r,echo=FALSE}
eigen_plot(img2)
```

**The eigenvector plots seem to almost mimic the original image, just in the opposite direction. As the index increases,the features contribute more. For the PC1 scores, each index represents a feature and how far it extends in the new direction.**

##Image 4

**Here are the rank approximations for the third image, as well as the plot of error terms and k values.**

```{r, echo=FALSE}
img4 <- read_csv("image4.csv")
img_iterations(img4)
```

**I believe that k = 10 is the most accurate approximation to the original image, even though there is certainly a noticeable difference between the two of them.**

**The plots of the PC1 scores and the eigenvectors will be listed below.**

```{r, echo=FALSE}
eigen_plot(img4)
```

**Compared to the other images, the eigenvector and PC1 scores look different. Unlike the other images, the PC1 score plot almost models a smooth function, meaning features follow a recognizable pattern. This is the case with the eigenvector plot as well, as where the larger values that drive the component are similar to each other in index.**

**This concludes the presentation.**
Thank you
-Elijah Echevarria
